---
title: "Using weave with your own data"
output:
  rmarkdown::html_vignette:
    css: weave.css
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Using weave with your own data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
source("_common.R")
set.seed(20260707)
```

```{r setup}
library(weave)
```

The [walkthrough](walkthrough.html) works on simulated data that arrives in
exactly the shape the model functions expect. Real data rarely does. This
article covers the practical on-ramp: preparing a raw data frame with
`data_process()`, encoding time correctly, what the coordinates mean, and how
to predict beyond the weeks you observed.

## 1. From a raw data frame to model-ready inputs

`infer_kernel_params()` and `gp_predict()` need two objects:

* `obs_data` — one row per site-week, with a factor `id`, a numeric time
  `t`, and the count `y_obs` (`NA` where missing);
* `coordinates` — one row per site, with `id`, `lon`, `lat`.

`data_process()` builds both from a single raw data frame. It expects columns
`t` (time), `n` (the count), `lat` and `lon`, plus any columns that together
identify a site — passed as bare names:

```{r raw-data}
# a raw extract: two regions, facilities reporting weekly counts, some weeks
# absent entirely (never recorded) and note facility C only starts at week 4
raw <- data.frame(
  region   = rep(c("North", "North", "South"), each = 6),
  facility = rep(c("A", "B", "C"), each = 6),
  t        = c(1:6,  1:6,  4:9),
  n        = rpois(18, 20),
  lat      = rep(c(0.10, 0.15, 0.90), each = 6),
  lon      = rep(c(0.20, 0.25, 0.80), each = 6)
)
raw <- raw[-c(3, 9), ]   # drop two site-weeks entirely, as real data does

prepared <- data_process(raw, region, facility)
str(prepared, max.level = 1)
```

Three things happened:

```{r prepared}
head(prepared$obs_data)
prepared$coordinates
prepared$nt
```

* the site-by-time grid was **completed** — every site now has a row for
  every time point that appears *anywhere* in the data, with `y_obs = NA`
  where nothing was recorded (absent rows and `NA` counts are treated
  identically: as missing). A week recorded by *no* site at all — week 3
  above — simply stays out of the grid: the kernels measure real elapsed
  time from `t`, so it is still modelled as a genuine gap;
* each site got a factor `id` (the original identifying columns are kept);
* sites that cannot be modelled — all counts missing, or missing
  coordinates — are dropped, with a message telling you which
  (`drop_zero = TRUE` also drops sites whose counts are all zero).

The returned pieces feed straight into the two model calls:

```{r fit, eval = FALSE}
est <- infer_kernel_params(prepared$obs_data, prepared$coordinates,
                           nt = prepared$nt, period = 52, refine = TRUE)
pred <- gp_predict(prepared$obs_data, prepared$coordinates,
                   hyperparameters = est, nt = prepared$nt, period = 52)
```

## 2. Encoding time correctly

`t` is not just a label: its *differences* are the distances the temporal
kernel sees. Three rules keep it honest:

* **Use a numeric index whose steps mirror real elapsed time** — e.g. weeks
  since the first week of the series (`1, 2, 3, ...` for consecutive weeks),
  or days since a reference date. A fortnight-long gap between two reports
  should show up as a difference of 2 weeks, not 1 row.
* **Never use an index that resets**, like week-of-year: week 52 of one year
  and week 1 of the next would look 51 weeks apart instead of adjacent.
* **`period` is in the same units as `t`** — weekly data with a yearly
  season means `period = 52`; daily data means `period = 365.25`.

Uneven spacing is fine — monthly data, or a mix of reporting frequencies,
just needs `t` values whose gaps reflect the real calendar. Both
`infer_kernel_params()` and `gp_predict()` must be given the *same* `t`
encoding, since the hyperparameters are estimated on that axis.

## 3. What the coordinates mean

Spatial distance is plain **Euclidean distance in the units of `lon`/`lat`**
— there is no great-circle correction. Two consequences:

* the fitted `length_scale` is in those units (degrees, if you pass raw
  longitude/latitude);
* one degree of longitude shrinks with latitude, so for a large study area
  or one far from the equator, **project the coordinates first** (e.g. to
  kilometres with a suitable local projection) and interpret `length_scale`
  in km.

For the small, near-equatorial extents typical of a single country or
region, raw degrees are usually an acceptable approximation.

## 4. Predicting beyond the observed weeks

`gp_predict()` predicts at every `(id, t)` cell present in `obs_data`. To
get predictions for weeks with no data anywhere — filling the future, or a
gap before the series starts — append rows with `NA` counts for those weeks
(for every site) and increase `nt` to match:

```{r forecast, eval = FALSE}
obs <- prepared$obs_data
future_weeks <- (max(obs$t) + 1):(max(obs$t) + 8)      # 8 weeks ahead
extra <- expand.grid(id = levels(obs$id), t = future_weeks)
extra$y_obs <- NA

obs_extended <- rbind(obs[, c("id", "t", "y_obs")], extra)
nt_extended  <- length(unique(obs_extended$t))

pred <- gp_predict(obs_extended, prepared$coordinates, hyperparameters = est,
                   nt = nt_extended, period = 52)
```

The prediction interval widens with distance from the data, as it should.
Treat genuine extrapolation with the usual caution: beyond the observed
range the GP leans increasingly on the fitted seasonal shape and long-term
drift rather than on evidence.

## 5. Scaling up

* **Estimation** costs \( O(n^3 + n_t^3) \) in the site and time counts. For
  very large site counts, pass `n_sites` to `infer_kernel_params()`: the
  hyperparameters are shared, population-level quantities, so a random site
  subsample estimates the same values at a fraction of the cost (set a seed
  first for reproducibility; time points are never subsampled).
* **Prediction** cost is dominated by the interval draws, which parallelise
  across CPU cores with one line of setup — see
  [*Running predictions in parallel*](parallel.html).
