---
title: "How weave fits the model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How weave fits the model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 8, fig.height = 5
)
```

This vignette walks through *how* `weave` fits its spatio-temporal model.
The goal is to give you a working mental picture of every block of the
sampler, without skipping the parts that make it correct. It assumes you
are comfortable with vectors, matrices, and a small amount of probability;
it does **not** assume any background in MCMC, Gaussian processes, or
Bayesian computation.

We'll cover, in order:

1. The model and what we want from it.
2. Why straightforward approaches don't work, and what we do instead.
3. The Pólya-Gamma identity --- the one trick that makes the rest easy.
4. The five blocks of one sampling sweep, explained one at a time.
5. Multiple chains, diagnostics, and what the output actually means.

---

## 1. The model

We have weekly case counts $y_{st}$ at health facility $s$ in week $t$.
We want to do three things with them at once:

- **Smooth over time**, so a single week's spike doesn't dominate.
- **Borrow strength across sites**, so a quiet facility benefits from its
  neighbours.
- **Fill in missing weeks** in a principled way, with honest uncertainty.

The model that does all three is:

$$
y_{st} \sim \mathrm{NegBin}(r, \mu = e^{\psi_{st}}), \qquad
\psi_{st} = \mu_s + f_{st},
$$

with the *latent field* $f$ given a Gaussian process prior:

$$
f \sim \mathcal{N}(0, \; K_{\text{space}} \otimes K_{\text{time}}).
$$

In plain English:

- Each count is **negative binomial** with mean $\lambda_{st} = e^{\psi_{st}}$
  and a dispersion parameter $r$ (small $r$ → heavy tails; $r \to \infty$
  → Poisson).
- The log-mean $\psi_{st}$ is a **site intercept** $\mu_s$ plus a smooth
  **latent surface** $f_{st}$.
- The surface $f$ has a Gaussian-process prior: it's smooth in space *and*
  in time, with the smoothness controlled by three knobs (the kernel
  hyperparameters $\theta$):
  - `length_scale` --- how quickly correlation decays with distance.
  - `periodic_scale` --- how peaky the seasonal cycle is.
  - `long_term_scale` --- how smooth the year-on-year trend is.

The covariance is **separable**: it factorises as a Kronecker product of a
small spatial kernel and a small temporal kernel. This is the structural
assumption that makes the maths tractable --- we never need to handle a
$(n \cdot n_t) \times (n \cdot n_t)$ matrix as a whole.

The parameters we want to learn from the data are $f$, $\mu_s$, $\theta$,
and $r$. "Learn" here means: get a sample of values that is representative
of the **posterior distribution** $p(f, \mu, \theta, r \mid y)$. Once we
have such a sample we can answer any question we want --- the posterior
mean for a missing week, a 95% credible interval, a forecast --- by
summarising the sample.

---

## 2. Why this is hard, and what we do instead

There are two reasons this is not a one-line maximum-likelihood problem.

**Reason A: the latent field is enormous.** For 1000 facilities and 150
weeks, $f$ has 150,000 entries. Standard tools like `Stan`'s NUTS can
sample over that many parameters in principle, but the geometry is awful
(the latent field is strongly correlated by construction), and you spend
most of your time tuning step sizes.

**Reason B: the likelihood is not Gaussian.** If $y_{st}$ were Gaussian, we
could write down the posterior of $f$ in closed form and be done. The
negative-binomial likelihood doesn't admit that, so we'd usually fall back
on a Laplace approximation. But Laplace can be badly biased when counts
are small, which is exactly the regime where routine HF data lives.

The strategy `weave` uses sidesteps both problems:

- Treat the model as a **Gibbs sampler**: cycle through the parameters,
  updating each block from its conditional posterior given everything
  else.
- Make every block either **exactly conjugate** (closed-form Gaussian or
  conjugate update) or a **simple one-dimensional sampler** (univariate
  slice, random-walk Metropolis).
- The hard block --- the 150,000-dim latent field $f$ --- becomes
  conjugate-Gaussian after a clever data augmentation (next section), so
  it's just one matrix-free linear solve per sweep.

The result is a sampler with no hand-tuning, exact (no Laplace anywhere),
and laptop-feasible for problem sizes in the thousands of sites and
hundreds of weeks.

---

## 3. The Pólya-Gamma identity

This is the only piece of maths you need to keep in mind. Everything else
follows from it.

The negative-binomial density, written with the linear predictor
$\eta = \psi - \log r$, has an awkward denominator:

$$
P(y \mid r, \eta) \;\propto\; \frac{e^{y \eta}}{(1 + e^{\eta})^{\,y + r}}.
$$

The denominator is what stops the conditional posterior of $\eta$ from
being Gaussian. The Pólya-Gamma identity (Polson, Scott, & Windle 2013)
says: there is an auxiliary random variable $\omega$ with the Pólya-Gamma
distribution such that

$$
\frac{e^{y \eta}}{(1 + e^{\eta})^{\,y + r}}
= \int_0^\infty \exp\!\left(\kappa \eta - \tfrac{\omega}{2}\eta^2\right)\,
  p_{\mathrm{PG}}(\omega \mid y + r, 0)\, d\omega,
\qquad \kappa = \frac{y - r}{2}.
$$

Read the integrand on the right as a function of $\eta$. With $\omega$
held fixed it is **exactly a Gaussian in $\eta$**, with mean $\kappa/\omega$
and precision $\omega$. The negative-binomial likelihood has been
re-expressed as a *Gaussian observation of $\eta$*, conditional on a
latent variable $\omega$ we can sample efficiently.

Practical consequence: introducing one $\omega$ per observed cell turns
the latent-field block from "non-Gaussian likelihood, no closed form" into
"Gaussian likelihood, closed-form conditional". The whole rest of the
sampler is built on this.

The `BayesLogit::rpg` function provides fast vectorised draws from the
Pólya-Gamma distribution.

---

## 4. Inside one sweep

A single sweep of the sampler updates every parameter once, in the
following order:

1. $\omega$ given $(f, \mu, \theta, r, y)$
2. $f$ given $(\omega, \mu, \theta, r, y)$
3. $\mu$ given $(\omega, f, r, y)$
4. $\theta$ given $f$
5. $r$ given $(y, \psi)$

We'll go through each block, explaining both the *what* (the conditional
distribution) and the *how* (the algorithm).

### 4.1 Block 1: $\omega \mid \mathrm{rest}$ --- the auxiliary refresh

For each observed cell $i$ we compute
$\eta_i = \psi_i - \log r = f_i + \mu_{s(i)} - \log r$, and draw

$$
\omega_i \sim \mathrm{PG}(y_i + r,\; \eta_i).
$$

This is one call to `BayesLogit::rpg` per sweep, vectorised over all
observed cells. It is cheap (microseconds per draw) and trivially
parallel.

**Intuition**: $\omega_i$ is the "weight" the Gaussian-disguised
likelihood will give to cell $i$ in the next block. Cells where the
likelihood is sharp get large $\omega$; cells where the likelihood is
flat get small $\omega$.

### 4.2 Block 2: $f \mid \mathrm{rest}$ --- the heavy block

After conditioning on $\omega$, the latent field has a Gaussian
conditional posterior. Writing $S$ for the matrix that selects the
observed cells out of the full $n \times n_t$ grid, and $\Omega =
\mathrm{diag}(\omega)$:

$$
\Sigma_{\text{post}}^{-1} = K^{-1} + S^{\!\top} \Omega S, \qquad
\mu_{\text{post}} = \Sigma_{\text{post}}\, S^{\!\top} \Omega\, (y^* - \mu),
$$

where $y^*_i = \kappa_i / \omega_i + \log r$ is the **pseudo-observation**
the PG identity gave us.

We never form $\Sigma_{\text{post}}$ --- it's a 150,000-dimensional dense
matrix at production scale. Instead we use the **perturbation sampler** of
Papandreou & Yuille (2010) / Bhattacharya et al. (2016), which produces an
exact draw from the conditional Gaussian with one matrix-free linear solve:

1. Sample a **prior draw** $u \sim \mathcal{N}(0, K)$. Because $K$ is
   Kronecker-separable, this is cheap (the existing `quick_mvnorm` does it
   in two Cholesky factorisations of small $n \times n$ and $n_t \times
   n_t$ matrices).
2. Sample an **observation-noise draw** $e_i \sim \mathcal{N}(0,
   1/\omega_i)$ at each observed cell.
3. Solve, by **preconditioned conjugate gradient (PCG)**, the small
   observation-space system
   $$(K + S^{\!\top}\Omega^{-1}S)\,\gamma = (y^* - \mu) - S u - e.$$
4. Return $f = u + K\, S^{\!\top}\gamma$.

The arithmetic to show this $f$ has exactly the right Gaussian
distribution is a few lines of algebra. The intuition: $u$ is the right
covariance but the wrong mean; the PCG solve is the precise correction
that moves it to the posterior mean while *preserving* the covariance.

**Why PCG is fast here.** PCG only needs two things: a way to multiply by
the system matrix, and a "preconditioner" --- a fast approximation to its
inverse. The system matrix never needs to exist as a matrix; the multiply
is a sequence of small operations using the Kronecker structure. The
preconditioner uses the eigendecomposition of the small $K_{\text{space}}$
and $K_{\text{time}}$ kernels --- also cheap. Typical convergence is in
50–150 iterations even on the production-scale problem.

### 4.3 Block 3: $\mu_s \mid \mathrm{rest}$ --- per-site Gaussian

The pseudo-observation $y^*_i$ is a Gaussian observation of $\psi_i = f_i
+ \mu_{s(i)}$ with precision $\omega_i$. Subtracting $f_i$ gives a
Gaussian observation of $\mu_{s(i)}$ alone. Combined with the
$\mathcal{N}(m_0, v_0)$ prior on $\mu_s$, the conditional posterior is
closed-form Normal per site:

$$
\text{prec} = \frac{1}{v_0} + \sum_{i\, \in\, \text{site } s} \omega_i,
\qquad
\text{mean} = \frac{m_0/v_0 + \sum_{i\, \in\, \text{site } s} \omega_i (y^*_i - f_i)}{\text{prec}}.
$$

Implemented as two `tapply` accumulations over sites and one vector of
`rnorm` draws --- fast and trivially correct.

### 4.4 Block 4: $\theta \mid f$ --- slice on the kernel hyperparameters

The three kernel hyperparameters $\theta = (\ell_s, \ell_p, \ell_t)$ each
get a **univariate slice sampler** on the log scale. The target is the
GP prior log-density of $f$ at the current value:

$$
\log p(\theta \mid f) = \log p(\theta) - \tfrac{1}{2}\log|K(\theta)|
- \tfrac{1}{2}\, f^{\!\top} K(\theta)^{-1} f.
$$

We can compute both $\log|K|$ and $f^{\!\top} K^{-1} f$ cheaply because of
the Kronecker structure: eigendecompose the small $K_{\text{space}}$ and
$K_{\text{time}}$ separately, and the log-determinant is a sum of the log
eigenvalues, while the quadratic form is a sum of squared entries of the
rotated $f$ divided by the eigenvalues.

The slice sampler itself (Neal 2003) needs no tuning. It picks a
horizontal "slice" at a uniform height under the log-target, finds an
interval bracketing the current value where the target is above the
slice, and then samples uniformly from that interval (shrinking on
rejection). Working on $\log \theta$ gives well-behaved stepping-out and
keeps every value positive.

This is the slowest block. Each accepted slice move requires an
$O(n^3)$ eigendecomposition of the spatial kernel, which is a few
seconds at $n = 1000$.

### 4.5 Block 5: $r \mid \mathrm{rest}$ --- random-walk Metropolis

The NB dispersion $r$ has no conjugate update, but it's a single positive
scalar and the posterior is well-behaved. We use a **random-walk
Metropolis-Hastings** update on $\log r$:

1. Propose $\log r' = \log r + \mathcal{N}(0, s^2)$.
2. Accept with probability
   $$\min\!\left\{1,\; \frac{p(y \mid \psi, r') p(r')}{p(y \mid \psi, r) p(r)} \cdot \frac{r'}{r}\right\}.$$
   (The trailing factor is the Jacobian for working on the log scale.)

The proposal scale $s$ is tuned adaptively during burn-in using
Robbins-Monro: nudge $\log s$ in the direction that brings the acceptance
rate toward the gold-standard 0.44 for 1-D random walks.

This is the only block with any "tuning", and we hide it inside burn-in.

---

## 5. Multiple chains, diagnostics, and posterior output

A single chain is enough for point estimates but not enough to be
*confident*. To detect non-convergence we run multiple chains in parallel
with deliberately different starting points, and compare them.

### 5.1 Overdispersed initial values

With `n_chains > 1`, `fit_bayes()` starts each chain after the first at
random draws from the priors of $\theta$ and $r$. The first chain still
uses sensible deterministic starts so it converges quickly. The other
chains are "overdispersed" --- different enough that if they all collapse
onto the same posterior region, we can trust the result.

### 5.2 Rhat (Gelman-Rubin diagnostic)

For each parameter we compute the **potential scale reduction factor**
$\hat R$. Informally:

$$
\hat R \approx \sqrt{\frac{\text{total variance across all chains}}{\text{average within-chain variance}}}.
$$

If the chains have all found the same distribution, the numerator and
denominator are equal and $\hat R = 1$. Values bigger than $\sim 1.05$ are
a sign chains haven't merged --- either the sampler is broken, the
posterior is multimodal, or the chains need more sweeps.

`summary.weave_bayes()` reports $\hat R$ for `length_scale`,
`periodic_scale`, `long_term_scale`, and `r` automatically when `n_chains
> 1`.

### 5.3 Effective sample size (ESS)

Chains are autocorrelated, so 1000 sweeps don't give 1000 independent
posterior draws. The effective sample size is the number of *equivalent
independent* draws. Aim for ESS $\geq 400$ per parameter for anything you
want to quote a credible interval on; $\geq 100$ for exploration.
`coda::effectiveSize` computes this from the trace.

### 5.4 Storage of the latent field

Storing $f$ at every sweep is a memory bomb (150k × 5000 sweeps × 3 chains
≈ 18 GB at 8 bytes/double). By default `fit_bayes()` keeps only a
**running mean and variance** of $f$ (Welford's algorithm; numerically
stable and $O(N)$ per update). For posterior-predictive draws you can
opt in to keeping a *thinned* set of full $f$ vectors --- the default of
100 evenly-spaced samples per chain is plenty for downstream plotting.

---

## 6. Putting it together: a worked example

Below is the minimal call that exercises everything we just discussed.
For more depth see the full walkthrough at
`implementation/test_walkthrough.R`.

```{r, eval = FALSE}
library(weave)
source(system.file("../implementation/simulation.R", package = "weave"))

# Simulate some HF count data.
set.seed(42)
n  <- 30; nt <- 104
coords <- data.frame(
  id  = factor(1:n),
  lat = runif(n, 0, 5),
  lon = runif(n, 0, 5),
  mu  = log(runif(n, 10, 80))
)
space_k <- space_kernel(coords, length_scale = 2)
time_k  <- time_kernel(1:nt, periodic_scale = 1, long_term_scale = 200)
true    <- simulate_data(n, nt, coords, space_k, time_k, r = 15)
obs     <- observed_data(true, p_one = 0.2, p_switch = 0.3)

# Fit.
fit <- fit_bayes(
  obs,
  n_sweeps = 1000, burnin = 400,
  n_chains = 3,
  store_f  = "thin"
)

# Diagnostics + posterior summary with Rhat.
summary(fit)
```

What to look at after the fit:

1. **Print the object**: chain count, sweeps, observation count, median
   PCG iterations per sweep, MH acceptance for $r$ per chain.
2. **Summary**: posterior medians and 95% intervals for the kernel
   parameters and $r$, plus $\hat R$.
3. **Posterior predictive**: `posterior_predict(fit)` gives one count-scale
   draw per stored MCMC sample at every cell. Coverage on held-out cells
   should be close to nominal.
4. **Run multiple chains**: the only honest way to know the sampler has
   converged is to see chains overlap and $\hat R$ near 1.

---

## 7. When the sampler is unhappy

A short field guide:

- **$\hat R > 1.1$** on `long_term_scale` after a moderate run: usually
  just slow mixing on a weakly identified parameter. Add more sweeps
  before worrying about model misspecification.
- **MH acceptance for $r$ outside $[0.15, 0.6]$** after burn-in: the
  adaptive scale didn't converge. Re-run with a longer burn-in.
- **PCG iteration count climbs across sweeps**: this can happen when
  $\omega$ becomes very heterogeneous (e.g. small $r$ + lots of zero
  cells). Increase `pcg_maxit` or relax `pcg_tol`.
- **Posterior predictive coverage well under 95%**: the model is mis-
  specified. The two usual culprits are (a) site-level deviation from
  the GP smoothness assumption --- some sites really are different ---
  and (b) the wrong period (e.g. you used `period = 52` for monthly
  data).

---

## 8. Further reading

- Polson, N., Scott, J., & Windle, J. (2013). *Bayesian inference for
  logistic models using Pólya-Gamma latent variables.* JASA 108(504).
- Bhattacharya, A., Chakraborty, A., & Mallick, B. K. (2016). *Fast
  sampling with Gaussian scale-mixture priors in high-dimensional
  regression.* Biometrika 103(4).
- Papandreou, G., & Yuille, A. (2010). *Gaussian sampling by local
  perturbations.* NeurIPS.
- Neal, R. (2003). *Slice sampling.* Annals of Statistics 31(3).
- Gelman, A., & Rubin, D. (1992). *Inference from iterative simulation
  using multiple sequences.* Statistical Science 7(4).
