---
title: "Estimating the kernel hyperparameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating the kernel hyperparameters}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  warning = FALSE,
  message = FALSE
)
set.seed(20260610)
```

```{r setup}
library(weave)
library(ggplot2)
```

This vignette is a full walkthrough of how `weave` turns noisy, gappy
count data into a smooth estimate of the underlying rate. It has two
halves:

1. **Estimate the kernel hyperparameters** — the handful of numbers that
   say how quickly counts become uncorrelated as you move apart in space
   and in time.
2. **Use them to predict** — denoise the observed counts, *fill in the
   missing weeks*, and put an honest **prediction interval** around each
   estimate.

We use a small simulated example with a *known* answer (including known
missing weeks) so you can see the method recover the truth.

At each stage we give: the **maths**, a **plain-language** explanation,
and the **worked example**.

The method is deliberately *quick*: it is a fast, deterministic estimate,
not a full Bayesian fit. The approximations that buy that speed are listed
honestly at the end.

---

# 1. The model

### The maths

We observe a count `y` at every site `s` (a health facility) and time `t`
(a week). We assume

\[
y_{st} \sim \text{NegBin}\!\left(r,\ \mu_{st} = e^{\,\mu_s + f_{st}}\right),
\qquad
f \sim \mathcal{N}\!\left(0,\ K_{\text{space}} \otimes K_{\text{time}}\right).
\]

- \( \mu_s \) is a per-site baseline (some facilities are just busier).
- \( f_{st} \) is a smooth **latent field**: the bit that wiggles up and
  down in space and time.
- \( K_{\text{space}} \) and \( K_{\text{time}} \) are **kernels** — they
  set how correlated two cells are as a function of how far apart they are.
- \( \otimes \) is the Kronecker product: the full space-time covariance is
  built from a small spatial kernel and a small temporal kernel
  ("separable").

The kernels have three hyperparameters — the knobs we want to estimate:

- `length_scale` — how far apart two sites have to be before they stop
  looking alike (spatial RBF kernel).
- `periodic_scale` — how "peaky" the yearly season is.
- `long_term_scale` — how slowly the year-on-year level drifts.

### In plain language

Each facility has a smooth underlying rate of cases. Nearby facilities,
and nearby weeks, tend to look similar; far-apart ones drift apart. The
three knobs control *how quickly* that "looking similar" fades with
distance in space and in time. Our whole job is to read those three
numbers off the data.

### The worked example

We simulate data where we *choose* the true knobs, so we can check the
answer. (`simulate_data()` and `observed_data()` below are small helpers
for the example — they are not part of the package.)

```{r}
# --- example controls --------------------------------------------------------
n      <- 30          # sites
nt     <- 52 * 3      # weeks (3 years)
period <- 52          # weeks per seasonal cycle

true_length_scale    <- 2     # spatial smoothness
true_periodic_scale  <- 1.1   # how peaky the season is
true_long_term_scale <- 150   # long-run drift
true_r               <- 15    # NB dispersion

# --- small simulation helpers (for the example only) -------------------------
simulate_data <- function(n, nt, coordinates, space_k, time_k, r) {
  f      <- quick_mvnorm(space_k, time_k)        # latent field, time fastest
  lambda <- exp(rep(coordinates$mu, each = nt) + f)
  data.frame(
    id     = factor(rep(seq_len(n), each = nt), levels = seq_len(n)),
    t      = rep(seq_len(nt), times = n),
    lambda = lambda,
    y      = stats::rnbinom(n * nt, size = r, mu = lambda)
  )
}

# Missing weeks arrive in CLUSTERS (whole stretches gone), as in real routine
# data, rather than one-off cells: a 0/1 sequence that tends to stay put.
generate_clustered_binary <- function(n, p_one, p_switch) {
  out <- numeric(n)
  out[1] <- stats::rbinom(1, 1, p_one)
  for (i in 2:n) {
    out[i] <- if (stats::runif(1) < p_switch) stats::rbinom(1, 1, p_one) else out[i - 1]
  }
  out
}
observed_data <- function(data, p_one = 0.15, p_switch = 0.2) {
  y_obs <- data$y
  y_obs[generate_clustered_binary(nrow(data), p_one, p_switch) == 1] <- NA
  cbind(data["id"], data["t"], y_obs = y_obs)
}

# --- draw the data -----------------------------------------------------------
coordinates <- data.frame(
  id  = factor(1:n),
  lon = runif(n, 0, 5),
  lat = runif(n, 0, 5),
  mu  = log(runif(n, 10, 80))
)

space_k <- space_kernel(coordinates, length_scale = true_length_scale)
time_k  <- time_kernel(1:nt,
  periodic_scale  = true_periodic_scale,
  long_term_scale = true_long_term_scale,
  period          = period
)

true_data <- simulate_data(n, nt, coordinates, space_k, time_k, r = true_r)
obs_data  <- observed_data(true_data)

cat(sprintf("%.0f%% of weeks are missing\n", 100 * mean(is.na(obs_data$y_obs))))
```

Here are the first four sites. The **black line** is the true underlying
mean (\( \lambda = e^{\mu_s + f_{st}} \)); **grey points** are the counts we
actually observe; and **red points** are the *held-out truth* at the weeks
that went missing — what the model will have to reconstruct, but never
gets to see.

```{r fig.height = 4.5}
show     <- 1:4
df_truth <- transform(true_data, id = as.integer(id))
df_obs   <- transform(obs_data,  id = as.integer(id))
df_miss  <- merge(df_truth[, c("id", "t", "y")], df_obs, by = c("id", "t"))
df_miss  <- df_miss[is.na(df_miss$y_obs), ]          # held-out truth

ggplot() +
  geom_line(data = subset(df_truth, id %in% show), aes(t, lambda)) +
  geom_point(data = subset(df_obs, id %in% show), aes(t, y_obs),
             size = 0.7, colour = "grey40") +
  geom_point(data = subset(df_miss, id %in% show), aes(t, y),
             size = 0.8, colour = "red") +
  facet_wrap(~ id, scales = "free_y", labeller = label_both) +
  labs(x = "Week", y = "Cases",
       title = "Simulated counts (4 of the sites)",
       subtitle = "black = true mean, grey = observed, red = held-out truth") +
  theme_bw()
```

---

# 2. Step 1 — a quick "plug-in" latent field

### The maths

We cannot see the latent field \( f \) directly. Instead we build a cheap
stand-in \( g \) from the counts:

\[
g_{st} = \frac{\log(1 + y_{st}) - \bar{g}_s}{\text{sd}_s},
\]

i.e. take \( \log(1+y) \), then for each site subtract that site's mean
and divide by its standard deviation.

### In plain language

The logarithm turns the multiplicative "rate" scale into an additive one
that matches \( f \). Subtracting each site's average removes the baseline
\( \mu_s \) (which we don't care about here), and dividing by the spread
puts every site on the same footing so a single set of knobs applies to
all of them. It's noisy — but it's instant.

### The worked example

```{r}
g <- build_plugin_field(obs_data, n = n, nt = nt)   # length n * nt, time fastest
str(g)
```

For one site, the plug-in field tracks the true latent field, just with
observation noise on top:

```{r}
# reshape g to sites x times and compare site 1 to the truth
G        <- t(matrix(g, nrow = nt, ncol = n))        # n x nt
f_true_m <- t(matrix(log(true_data$lambda) - rep(coordinates$mu, each = nt),
                     nrow = nt, ncol = n))           # true f, centred per site

site <- 1
plot_df <- rbind(
  data.frame(t = 1:nt, value = scale(f_true_m[site, ])[, 1], type = "true f (scaled)"),
  data.frame(t = 1:nt, value = G[site, ],                    type = "plug-in g")
)
ggplot(plot_df, aes(t, value, colour = type)) +
  geom_line() +
  labs(x = "Week", y = "Standardised latent field", title = "Site 1: plug-in vs truth",
       colour = NULL) +
  theme_bw()
```

---

# 3. Step 2 — score a set of knobs with the marginal likelihood

### The maths

For a candidate set of hyperparameters \( \theta \) the kernels give a
covariance \( K(\theta) \). We add a **nugget** — a noise term \( \eta \)
— and a global variance \( \sigma^2 \):

\[
g \sim \mathcal{N}\!\Big(0,\ \sigma^2\big[(R_{\text{space}} \otimes R_{\text{time}}) + \eta I\big]\Big),
\]

where \( R \) are *correlation* kernels (unit diagonal). The score for
\( \theta \) is the log marginal likelihood

\[
\log p(g \mid \theta) = -\tfrac{1}{2}\Big(\log|K| + g^\top K^{-1} g + N\log 2\pi\Big).
\]

We pick the \( \theta \) that maximises this. The variance \( \sigma^2 \)
has a closed-form optimum, so we "profile" it out and never search over
it.

### In plain language

For any guess at the knobs, we can ask: *how plausible is the plug-in
field if the world really had this much smoothness?* The first term
(\( \log|K| \)) penalises over-flexible kernels (an "Occam" penalty); the
second rewards kernels that explain the field well. We turn the dial until
the score is best.

The **nugget** \( \eta \) is the important addition: it is a slice of
"pure noise" the model is allowed to attribute to observation error
instead of signal. Without it, the kernel would have to explain the noisy
plug-in field with smoothness alone, badly distorting the length scales.

---

# 4. Step 3 — why it is fast: the Kronecker trick

### The maths

The full covariance is \( (n \cdot n_t) \times (n \cdot n_t) \) — here
`r n * nt` square. We never form it. Because it is a Kronecker product,
eigendecomposing the *small* factors \( R_{\text{space}} = U_s
\Lambda_s U_s^\top \) (size \( n \)) and \( R_{\text{time}} = U_t
\Lambda_t U_t^\top \) (size \( n_t \)) is enough:

\[
\log|K + \eta I| = \sum_{i,j} \log(a_i b_j + \eta), \qquad
g^\top (K + \eta I)^{-1} g = \sum_{i,j} \frac{G_{ij}^2}{a_i b_j + \eta},
\]

with \( a_i, b_j \) the eigenvalues, \( G = U_s^\top F U_t \), and \( F \)
the field reshaped to \( n \times n_t \). The cost is
\( O(n^3 + n_t^3) \) instead of \( O\big((n\,n_t)^3\big) \).

### In plain language

A separable kernel means space and time can be handled separately. So
instead of one enormous matrix we juggle two small ones — one for space,
one for time — which is astronomically cheaper, and gives the *exact* same
answer. This is the whole reason the estimate takes a second rather than
hours.

---

# 5. Step 4 — fit, and check against the truth

### The worked example

`infer_kernel_params()` does all of the above: build the plug-in field,
then maximise the marginal likelihood over the three length scales plus
the nugget ratio.

```{r}
est <- infer_kernel_params(obs_data, coordinates, nt = nt, period = period)

data.frame(
  parameter = c("length_scale", "periodic_scale", "long_term_scale"),
  truth     = c(true_length_scale, true_periodic_scale, true_long_term_scale),
  estimate  = round(c(est$length_scale, est$periodic_scale, est$long_term_scale), 2)
)
```

```{r}
cat(sprintf("nugget ratio (noise/signal) = %.2f;  profiled sigma^2 = %.2f\n",
            est$nugget_ratio, est$sigma2))
```

The clearest check is to draw the **fitted** kernels (pink) on top of the
**true** ones (black dashed). If the method worked, they overlap.

```{r}
kernel_curves <- function(ls, ps, lts, label) {
  sp <- data.frame(distance = seq(0, 5, length.out = 200), type = label)
  sp$correlation <- rbf_kernel(sp$distance, theta = ls)
  tm <- data.frame(lag = seq(0, nt - 1, length.out = 400), type = label)
  tm$correlation <- periodic_kernel(tm$lag, alpha = ps, period = period) *
    rbf_kernel(tm$lag, theta = lts)
  list(space = sp, time = tm)
}
fit  <- kernel_curves(est$length_scale, est$periodic_scale, est$long_term_scale, "Estimated")
true <- kernel_curves(true_length_scale, true_periodic_scale, true_long_term_scale, "True")

cols <- c(Estimated = "deeppink", True = "black")
ltys <- c(Estimated = "solid",    True = "dashed")

ggplot(rbind(fit$space, true$space),
       aes(distance, correlation, colour = type, linetype = type)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = cols, name = NULL) +
  scale_linetype_manual(values = ltys, name = NULL) +
  labs(x = "Spatial distance", y = "Correlation",
       title = "Spatial kernel: estimated vs true") +
  ylim(0, 1) + theme_bw()

ggplot(rbind(fit$time, true$time),
       aes(lag, correlation, colour = type, linetype = type)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = cols, name = NULL) +
  scale_linetype_manual(values = ltys, name = NULL) +
  labs(x = "Temporal lag (weeks)", y = "Correlation",
       title = "Temporal kernel: estimated vs true") +
  theme_bw()
```

---

# 6. Using the estimate: smoothing, gap-filling, and prediction intervals

Now we put the fitted kernels to work: clean up the noise, fill in the
missing weeks, and say how *uncertain* each estimate is.

## 6.1 Smoothing the latent rate (the best guess)

### The maths

With the kernels fixed, denoising the plug-in field is a closed-form
Gaussian-process *smoother*. In the same Kronecker eigenbasis every mode
\( (i,j) \) is shrunk towards zero, and we also get its posterior variance:

\[
\widehat{G}_{ij} = \underbrace{\frac{a_i b_j}{a_i b_j + \eta}}_{\text{shrink } S_{ij}}\, G_{ij},
\qquad
\operatorname{Var}(\widehat{G}_{ij}) = \sigma^2\,\frac{a_i b_j\,\eta}{a_i b_j + \eta}.
\]

Strong-signal modes (large \( a_i b_j \)) are kept; noisy modes (small) are
damped. Undo the per-site standardisation and exponentiate to get the
posterior **mean** and **variance** of the log-rate at *every* cell —
missing ones included.

### In plain language

The fitted kernels tell us which wiggles are real signal and which are
noise. The smoother keeps the real ones and irons out the rest, borrowing
strength from neighbouring sites and weeks to fill the gaps. It also keeps
track of how confident it is at each point — wider where data is sparse or
missing.

### The worked example

```{r}
gp_smoother <- function(obs_data, coordinates, est, n, nt, period) {
  ids <- sort(unique(obs_data$id)); times <- sort(unique(obs_data$t))
  coordinates <- coordinates[match(ids, coordinates$id), , drop = FALSE]

  # plug-in field, kept un-standardised so we can put it back afterwards
  M <- matrix(NA_real_, n, nt)
  M[cbind(match(obs_data$id, ids), match(obs_data$t, times))] <- log1p(obs_data$y_obs)
  row_mean <- rowMeans(M, na.rm = TRUE); row_mean[!is.finite(row_mean)] <- 0
  Mc <- M - row_mean
  row_sd <- apply(Mc, 1, stats::sd, na.rm = TRUE)
  row_sd[!is.finite(row_sd) | row_sd == 0] <- 1
  G <- Mc / row_sd; G[is.na(G)] <- 0

  es <- eigen(space_kernel(coordinates, length_scale = est$length_scale), symmetric = TRUE)
  et <- eigen(time_kernel(times, periodic_scale = est$periodic_scale,
                          long_term_scale = est$long_term_scale, period = period),
              symmetric = TRUE)
  a <- pmax(es$values, 1e-12); b <- pmax(et$values, 1e-12)
  lam <- outer(a, b)

  shrink <- lam / (lam + est$nugget_ratio)                       # posterior mean
  pv     <- est$sigma2 * lam * est$nugget_ratio / (lam + est$nugget_ratio) # posterior var

  ghat <- crossprod(es$vectors, G) %*% et$vectors
  Ghat <- es$vectors %*% (shrink * ghat) %*% t(et$vectors)       # smoothed field
  Vstd <- (es$vectors^2) %*% pv %*% t(et$vectors^2)              # per-cell variance

  # undo standardisation -> posterior mean (Z) and variance (Vz) of the log-rate
  Z  <- row_mean + row_sd * Ghat
  Vz <- (row_sd^2) * Vstd
  list(Z = Z, Vz = Vz, mean = exp(Z + Vz / 2))                   # mean = E[lambda]
}

sm <- gp_smoother(obs_data, coordinates, est, n, nt, period)     # Z, Vz, mean (n x nt)
```

## 6.2 A prediction interval for the counts

### The maths

`sm` gives the rate \( \lambda \): a lognormal posterior with
\( \log\lambda \sim \mathcal{N}(Z, V_z) \). But an actual *count* also
bounces around its rate via the Negative-Binomial observation noise,
\( y \mid \lambda \sim \text{NegBin}(r, \lambda) \). We combine the two with
the **law of total variance**:

\[
\mathbb{E}[y] = \mathbb{E}[\lambda], \qquad
\operatorname{Var}[y] = \underbrace{\mathbb{E}\!\big[\lambda + \tfrac{\lambda^2}{r}\big]}_{\text{average count noise}}
                       + \underbrace{\operatorname{Var}[\lambda]}_{\text{rate uncertainty}} .
\]

We then moment-match a lognormal to \( (\mathbb{E}[y], \operatorname{Var}[y]) \)
and read off the 2.5% and 97.5% points. The dispersion \( r \) is the one
thing this quick method does not estimate, so we recover it by method of
moments from the observed cells (using \( \operatorname{Var}(y\mid\lambda)
= \lambda + \lambda^2/r \)).

### In plain language

There are two reasons a future count is uncertain: we are not sure of the
underlying rate (the smoother's wider/narrower confidence), **and** counts
scatter around any given rate (more so when the data is overdispersed). A
genuine prediction interval has to include both. We add them up and draw a
band that should contain about 95% of real counts. The "how overdispersed"
number \( r \) is read straight off the data.

### The worked example

```{r}
# 1. estimate the NB dispersion r by method of moments (observed cells only)
ids <- sort(unique(obs_data$id)); times <- sort(unique(obs_data$t))
lam_at <- sm$mean[cbind(match(obs_data$id, ids), match(obs_data$t, times))]
keep   <- !is.na(obs_data$y_obs)
yk <- obs_data$y_obs[keep]; lk <- lam_at[keep]
den   <- sum((yk - lk)^2 - lk)
r_hat <- if (den > 0) max(sum(lk^2) / den, 0.1) else 1e6          # large r -> ~Poisson
cat(sprintf("estimated dispersion r = %.1f  (true = %.0f)\n", r_hat, true_r))

# 2. predictive mean & variance of counts, then lognormal 2.5/97.5% bounds
Elam  <- exp(sm$Z + sm$Vz / 2)
Elam2 <- exp(2 * sm$Z + 2 * sm$Vz)
Vlam  <- exp(2 * sm$Z + sm$Vz) * (exp(sm$Vz) - 1)
pmean <- Elam
pvar  <- Elam + Elam2 / r_hat + Vlam
ss    <- log(1 + pvar / pmax(pmean, 1e-8)^2)
mln   <- log(pmax(pmean, 1e-8)) - ss / 2

pred_df <- data.frame(
  id    = rep(seq_len(n), each = nt),       # site-major, time-fastest --
  t     = rep(seq_len(nt), times = n),      # matches as.vector(t(.)) below
  mean  = as.vector(t(pmean)),
  lower = as.vector(t(stats::qlnorm(0.025, mln, sqrt(ss)))),
  upper = as.vector(t(stats::qlnorm(0.975, mln, sqrt(ss))))
)
```

The blue line is the predicted mean and the blue band the 95% prediction
interval. It should track the black true-mean line and cover the points —
including the red held-out weeks it never saw.

```{r fig.height = 4.5}
ggplot() +
  geom_ribbon(data = subset(pred_df, id %in% show),
              aes(t, ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(data = subset(pred_df, id %in% show), aes(t, mean),
            colour = "steelblue", linewidth = 0.7) +
  geom_line(data = subset(df_truth, id %in% show), aes(t, lambda),
            colour = "black", linewidth = 0.4) +
  geom_point(data = subset(df_obs, id %in% show), aes(t, y_obs),
             size = 0.7, colour = "grey40") +
  geom_point(data = subset(df_miss, id %in% show), aes(t, y),
             size = 0.8, colour = "red") +
  facet_wrap(~ id, scales = "free_y", labeller = label_both) +
  labs(x = "Week", y = "Cases",
       title = "Prediction vs truth",
       subtitle = "blue = predicted mean + 95% interval; black = true mean; red = held-out truth") +
  theme_bw()
```

## 6.3 Did we fill the gaps well?

### The worked example

The honest test is the **held-out** weeks: does the 95% interval contain
the counts the model never saw? It should, about 95% of the time.

```{r}
# held-out cells only, with the true rate (lambda) and held-out count (y)
chk <- merge(df_truth[, c("id", "t", "lambda", "y")], pred_df, by = c("id", "t"))
chk <- merge(chk, df_miss[, c("id", "t")], by = c("id", "t"))   # keep held-out cells

cat(sprintf("Held-out weeks: %d\n", nrow(chk)))
cat(sprintf("95%% interval covers the held-out count: %.0f%% of the time\n",
            100 * mean(chk$y >= chk$lower & chk$y <= chk$upper)))
cat(sprintf("correlation of predicted mean with true rate: %.2f\n",
            cor(chk$mean, chk$lambda)))
```

### In plain language

If the coverage lands near 95% the bands are honest; if the predicted mean
correlates strongly with the true rate, the gap-filling is doing its job.
Both can drift a little here because this is a small example and the
quick method does not propagate every source of uncertainty (see below).

---

# 7. Caveats and approximations

This is a fast, deterministic estimate — not a full posterior. The
shortcuts that make it quick also limit it:

- **Plug-in, not Bayesian.** We condition on a single noisy estimate of
  the latent field rather than integrating it out. This *attenuates* the
  length scales (`periodic_scale` and `long_term_scale` tend to come out a
  little low). The nugget mitigates this but does not remove it.
- **`log(1 + y)`** is a crude stand-in for the latent log-rate, and is
  poorest at very low counts.
- **Per-site scaling** folds all per-site variance into a single global
  \( \sigma^2 \) and amplifies noise at low-count sites.
- **One homoscedastic nugget.** Real count noise is heteroscedastic (it
  grows with the rate); we summarise it with a single ratio \( \eta \).
- **Missing cells are mean-imputed**, so they add no uncertainty of their
  own — the prediction interval can therefore be a little optimistic over
  long gaps.
- **The dispersion \( r \)** is recovered by a rough method of moments, so
  the *width* of the count prediction interval inherits that error.
- **A point estimate.** We report the maximiser, with no uncertainty on
  the hyperparameters themselves.

For a quick, honest summary of the spatial and temporal correlation
structure — or a sensible starting point for a fuller model — it does the
job in about a second.
```
