---
title: "The Kronecker trick: why weave is fast"
output:
  rmarkdown::html_vignette:
    css: weave.css
vignette: >
  %\VignetteIndexEntry{The Kronecker trick: why weave is fast}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
library(ggplot2)
source("_common.R")
```

```{r setup, include = FALSE}
library(weave)
```

A weave model has one cell per site and week: \( N = n \cdot n_t \) cells in
total. The Gaussian-process covariance over those cells is an
\( N \times N \) matrix — at 1000 sites and five years of weeks that is
around 67 *billion* entries. weave never forms it, never inverts it, and
never approximates it. Everything in
[the walkthrough](walkthrough.html) — fitting the hyperparameters and
computing the exact part of the prediction interval — stays exact because
the covariance is **separable**, and separable covariances can be handled
through their small factors alone. This article explains the two identities
that make that work.

## One small matrix per dimension

The model assumes the covariance factorises into a spatial part and a
temporal part:

\[
K \;=\; K_{\text{space}} \otimes K_{\text{time}},
\]

where \( K_{\text{space}} \) is \( n \times n \), \( K_{\text{time}} \) is
\( n_t \times n_t \), and \( \otimes \) is the **Kronecker product** — a
recipe that builds the one enormous matrix from the two small ones: the
covariance between cell \( (s, t) \) and cell \( (s', t') \) is simply
\( K_{\text{space}}[s, s'] \times K_{\text{time}}[t, t'] \).

The structure is easiest to see. Below is the full space-time covariance for
a toy problem of 3 sites × 8 weeks: a 24 × 24 matrix, but built entirely from
a 3 × 3 spatial kernel and an 8 × 8 temporal one. Each 8 × 8 block is the
temporal kernel scaled by one entry of the spatial kernel — the whole matrix
never needs to exist:

```{r kron-schematic, echo = FALSE, fig.height = 4.4, fig.alt = "Heatmap of a 24-by-24 Kronecker-product covariance matrix showing a 3-by-3 grid of 8-by-8 blocks; each block is the temporal kernel scaled by one spatial kernel entry"}
n_toy <- 3; nt_toy <- 8
coords_toy <- data.frame(id = 1:n_toy, lon = c(0, 1, 4), lat = c(0, 1, 0))
Ks_toy <- space_kernel(coords_toy, length_scale = 2)
Kt_toy <- time_kernel(1:nt_toy, periodic_scale = 1, long_term_scale = 50,
                      period = 8)
K_toy  <- kronecker(Ks_toy, Kt_toy)

kron_df <- expand.grid(row = 1:(n_toy * nt_toy), col = 1:(n_toy * nt_toy))
kron_df$value <- as.vector(K_toy)

ggplot(kron_df, aes(col, row, fill = value)) +
  geom_tile() +
  geom_hline(yintercept = c(8.5, 16.5), colour = "white", linewidth = 0.8) +
  geom_vline(xintercept = c(8.5, 16.5), colour = "white", linewidth = 0.8) +
  scale_y_reverse() +
  scale_fill_gradient(low = "#f7f1e8", high = weave_cols[["prediction"]],
                      name = "covariance") +
  coord_fixed() +
  labs(x = NULL, y = NULL,
       title = "The full covariance for 3 sites × 8 weeks",
       subtitle = "each 8×8 block = temporal kernel × one spatial kernel entry") +
  theme_weave() +
  theme(axis.text = element_blank(), panel.grid.major = element_blank())
```

Statistically, separability says the *shape* of the temporal correlation is
the same at every site (and vice versa) — a modelling assumption, and the
price of everything below.

## Identity 1: determinants and quadratic forms (used in fitting)

Fitting the hyperparameters requires the log marginal likelihood of the
plug-in field \( g \), which needs \( \log|K| \) and
\( g^\top K^{-1} g \) — normally hopeless at size \( N \). Because the
covariance is a Kronecker product plus a diagonal nugget, eigendecomposing
the *small* factors
\( R_{\text{space}} = U_s \Lambda_s U_s^\top \) (size \( n \)) and
\( R_{\text{time}} = U_t \Lambda_t U_t^\top \) (size \( n_t \)) is enough:

\[
\log\bigl|R_{\text{space}} \otimes R_{\text{time}} + \eta I\bigr|
  = \sum_{i,j} \log(a_i b_j + \eta), \qquad
g^\top \bigl(R_{\text{space}} \otimes R_{\text{time}} + \eta I\bigr)^{-1} g
  = \sum_{i,j} \frac{G_{ij}^2}{a_i b_j + \eta},
\]

with \( a_i, b_j \) the eigenvalues, \( G = U_s^\top F U_t \), and \( F \)
the field reshaped to \( n \times n_t \). This works because the Kronecker
product's eigenvectors are themselves a Kronecker product
(\( U_s \otimes U_t \)), and a nugget \( \eta I \) only shifts the
eigenvalues. The cost is \( O(n^3 + n_t^3) \) instead of
\( O\big((n\,n_t)^3\big) \) — for the 1000-site example above, that is the
difference between a second and centuries.

The same eigendecomposition gives the *exact* complete-grid posterior
variance used by the prediction interval — the
\( V_{\text{complete}} \) term in the walkthrough — again without ever
forming \( K \).

## Identity 2: multiplying by K without forming it (used in prediction)

Predicting needs solves of the form
\( (S K S^\top + \nu I)^{-1} b \) (one for the posterior mean, one per
perturbation draw). weave uses the conjugate-gradient method, which never
needs \( K \) itself — only the ability to compute \( K v \) for a given
vector \( v \). For a Kronecker product that is a reshape away:

\[
\bigl(K_{\text{space}} \otimes K_{\text{time}}\bigr)\, v
  \;=\; \operatorname{vec}\!\bigl(K_{\text{time}}\, V\, K_{\text{space}}\bigr),
\]

where \( V \) is \( v \) reshaped to an \( n_t \times n \) matrix (weave's
vector layout puts time fastest, and the kernels are symmetric). Each
product costs two small matrix multiplications instead of one enormous one —
this is the package's `kron_mv()` building block, and it is what makes each
conjugate-gradient iteration cheap even at large \( N \).

## What this buys

::: {.intuition}
A separable kernel means space and time can be handled separately. So
instead of one enormous matrix we juggle two small ones — one for space, one
for time — which is dramatically cheaper and gives the *exact* same answer.
This is why the estimate takes a second rather than hours.
:::

Everything downstream inherits the speed: hyperparameter fitting evaluates
the exact likelihood thousands of times inside the optimiser
(identity 1), the exact part of the prediction interval is closed-form
(identity 1 again), and the gap-correction draws each cost a handful of
cheap matrix-vector products (identity 2). None of it is an approximation —
the only approximations in weave are the statistical ones listed at the end
of [the walkthrough](walkthrough.html).
