---
title: "Running predictions in parallel"
output:
  rmarkdown::html_vignette:
    css: weave.css
vignette: >
  %\VignetteIndexEntry{Running predictions in parallel}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
source("_common.R")
```

This short article continues the example from
[the walkthrough](walkthrough.html) — `obs_data` and `est` below are the
data and fitted hyperparameters built there.

The expensive part of `gp_predict()` is the set of `n_draws` simulation
draws that estimate how the missing weeks widen the prediction interval.
Each draw is an independent solve, so at scale (many sites, many draws) they
parallelise naturally across CPU cores.

`weave` uses the [future](https://future.futureverse.org/) framework for
this. By default everything runs serially — nothing to install, nothing to
configure — and you opt in with a single line before the call. No arguments
to `gp_predict()` change.

```{r parallel, eval = FALSE}
future::plan(future::multisession, workers = 4)   # 4 background R sessions

pred <- gp_predict(obs_data, coordinates, hyperparameters = est,
                   nt = nt, period = period, n_draws = 100)

future::plan(future::sequential)                  # back to serial
```

`future::multisession` works on every platform (it uses background R
sessions, not forks). Any other future backend — `multicore` on Linux/macOS,
a cluster of machines — works the same way; see the
[future documentation](https://future.futureverse.org/) for the options.

## The numbers never change

Parallelism is purely a speed decision:

* **Results are identical** for every backend and worker count. Each draw
  gets its own pre-generated random-number stream, so the split across
  workers cannot affect the numbers.
* **Reproducibility** works exactly as in serial code: `set.seed(1)` before
  `gp_predict()` gives the same answer whether you run on 1 worker or 8.

## When it pays off

Each worker takes about a second to start and receives the kernel matrices
once. Parallelism therefore helps when the number of sites × draws is large;
for small problems the serial default is already the fastest option. As an indication:
a 300-site × 260-week problem with 10% missing data and 100 draws ran about
4× faster with `workers = 4` on a 4-performance-core laptop.

## Practicalities

* The **progress bar** only shows when running serially — background workers
  cannot draw it.
* At very large site counts you may need to raise the future framework's
  export cap: `options(future.globals.maxSize = 1e9)` (the matrices shipped
  to each worker are about 40 MB at 1000 sites × 260 weeks).
* If your R uses a **multithreaded BLAS** (OpenBLAS, MKL), cap its threads
  inside workers to avoid oversubscription. R's shipped BLAS is
  single-threaded, so by default there is nothing to do.
