---
title: "The weave walkthrough: estimating kernels and predicting rates"
output:
  rmarkdown::html_vignette:
    css: weave.css
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{The weave walkthrough: estimating kernels and predicting rates}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
library(ggplot2)
source("_common.R")
set.seed(20260610)
```

```{r setup}
library(weave)
```

This article is a complete walkthrough of how `weave` turns noisy, gappy count
data into a smooth estimate of an underlying rate. If the ideas of a Gaussian
process and a kernel are new, the companion
[*gentle introduction*](gaussian-processes.html) builds them up first; here we
apply them. The walkthrough has three parts:

1. **The model** — what we assume generates the counts, the kernels that
   encode smoothness, and the simulated data we will use throughout.
2. **Fitting the kernel hyperparameters** — how the parameters that control
   how quickly counts become uncorrelated in space and time are estimated,
   including the EM-style handling of missing weeks.
3. **From fitted kernels to predictions** — the best-guess rate at every
   site and week, and an honest **prediction interval** around it, with
   `gp_predict()`.

We use a small simulated example with a *known* answer (including known missing
weeks), so we can watch the method recover the truth. At each stage we give the
**maths**, a **plain-language** explanation, and the **worked example**.

::: {.caveat}
The method is deliberately *quick*: a fast, deterministic estimate rather than
a full Bayesian fit. The approximations that buy that speed are listed honestly
at the end.
:::

## 1. The model

### The observation model: counts by site and week

::: {.maths}
We observe a count `y` at every site `s` (a health facility) and time `t`
(a week), and assume

\[
y_{st} \sim \text{NegBin}\!\left(r,\ \mu_{st} = e^{\,\mu_s + f_{st}}\right),
\qquad
f \sim \mathcal{N}\!\left(0,\ K_{\text{space}} \otimes K_{\text{time}}\right).
\]

- \( \mu_s \) is a per-site baseline (some facilities are simply busier).
- \( f_{st} \) is a smooth **latent field** — the part that varies up and down
  in space and time.
- \( K_{\text{space}} \) and \( K_{\text{time}} \) are **kernels**: they set
  how correlated two cells are as a function of their separation.
- \( \otimes \) is the Kronecker product — the full space-time covariance is
  built from a small spatial kernel and a small temporal kernel
  ("separable"). This structure is also why everything below is fast: see
  [*The Kronecker trick*](kronecker.html).

The Negative-Binomial assumption is lighter than it looks: it enters only the
**prediction interval** (through the count-noise variance
\( \lambda + \lambda^2/r \)). The hyperparameter estimates and the predicted
rate are built from the Gaussian field on the \( \log(1+y) \) scale and do
not depend on it. With no overdispersion the model reduces to Poisson —
the limit \( r = \infty \), which the dispersion estimate returns
automatically when the data show no excess variance.
:::

::: {.intuition}
Each facility has a smooth underlying rate of cases. Nearby facilities, and
nearby weeks, tend to look similar; far-apart ones drift. The kernels control
*how quickly* that resemblance fades with distance in space and in time. Our
whole task is to read their parameters off the data.
:::

### The kernels and their hyperparameters

The spatial kernel is an RBF (squared-exponential) on the distance between
sites; the temporal kernel is a periodic kernel (the seasonal cycle)
multiplied by a slowly-decaying RBF (the long-run drift). Between them they
carry three hyperparameters — the quantities we want to estimate:

- `length_scale` — how far apart two sites must be before they stop looking
  alike (spatial RBF kernel);
- `periodic_scale` — how sharply the yearly season rises and falls;
- `long_term_scale` — how slowly the year-on-year level drifts.

A fourth, the **nugget ratio**, is estimated alongside them: the share of the
variance that is observation noise rather than smooth signal. Its role is
explained in part 2. (The [*gentle introduction*](gaussian-processes.html)
shows what each kernel looks like and how its parameter changes the curves it
generates.)

### The worked example: simulated data with missing weeks

We simulate data with *chosen* true hyperparameters, so we can check the
answer. The small helpers that draw the counts and remove clustered runs of
weeks (`simulate_data()`, `observed_data()`) are example scaffolding, not part
of the package, so they are hidden here.

```{r sim-helpers, echo = FALSE}
# Example-only helpers (NOT part of weave). They draw a Negative-Binomial GP and
# then remove clustered runs of weeks, mimicking real reporting dropouts.
simulate_data <- function(n, nt, coordinates, space_k, time_k, r) {
  f      <- quick_mvnorm(space_k, time_k)        # latent field, time fastest
  lambda <- exp(rep(coordinates$mu, each = nt) + f)
  data.frame(
    id     = factor(rep(seq_len(n), each = nt), levels = seq_len(n)),
    t      = rep(seq_len(nt), times = n),
    lambda = lambda,
    y      = stats::rnbinom(n * nt, size = r, mu = lambda)
  )
}
generate_clustered_binary <- function(n, p_one, p_switch) {
  out <- numeric(n)
  out[1] <- stats::rbinom(1, 1, p_one)
  for (i in 2:n) {
    out[i] <- if (stats::runif(1) < p_switch) stats::rbinom(1, 1, p_one) else out[i - 1]
  }
  out
}
observed_data <- function(data, p_one = 0.15, p_switch = 0.2) {
  y_obs <- data$y
  y_obs[generate_clustered_binary(nrow(data), p_one, p_switch) == 1] <- NA
  cbind(data["id"], data["t"], y_obs = y_obs)
}
```

```{r controls}
# example dimensions
n      <- 20          # sites (health facilities)
nt     <- 52 * 3      # weeks (3 years)
period <- 52          # weeks per seasonal cycle

# the true kernel hyperparameters we will try to recover
true_length_scale    <- 2     # spatial smoothness
true_periodic_scale  <- 1.1   # how sharp the season is
true_long_term_scale <- 150   # long-run drift
true_r               <- 15    # Negative-Binomial dispersion

# site coordinates and per-site baseline log-rate (mu_s)
coordinates <- data.frame(
  id  = factor(1:n),
  lon = runif(n, 0, 5),
  lat = runif(n, 0, 5),
  mu  = log(runif(n, 10, 80))
)

# build the true kernels, draw counts, then hide clustered gaps.
# `obs_data` is all that weave sees: columns id, t, y_obs (NA = missing).
space_k   <- space_kernel(coordinates, length_scale = true_length_scale)
time_k    <- time_kernel(1:nt, periodic_scale = true_periodic_scale,
                         long_term_scale = true_long_term_scale, period = period)
true_data <- simulate_data(n, nt, coordinates, space_k, time_k, r = true_r)
obs_data  <- observed_data(true_data)

cat(sprintf("%.0f%% of weeks are missing\n", 100 * mean(is.na(obs_data$y_obs))))
```

::: {.caveat}
Two things the simulation glosses over for real data. First, `t` must be a
numeric index whose *differences* encode real elapsed time — e.g. weeks since
a reference date — and `period` must be in those same units; arbitrary
indices (week-of-year that resets, row numbers) silently distort every
temporal length-scale. Second, real data rarely arrives in this shape:
`data_process()` turns one raw data frame into the `obs_data`, `coordinates`
and `nt` used below — see
[*Using weave with your own data*](data-preparation.html).
:::

The gaps are not scattered at random: they arrive in *clustered runs*, the
way real reporting drops out — a facility goes quiet for a stretch, then
comes back. Magenta marks the missing weeks:

```{r missingness-map, echo = FALSE, fig.height = 3.8, fig.alt = "Heatmap of sites by weeks: observed weeks in grey, missing weeks in magenta, with the missing weeks forming horizontal clustered runs within sites"}
miss_map <- data.frame(
  id      = as.integer(obs_data$id),
  t       = obs_data$t,
  status  = ifelse(is.na(obs_data$y_obs), "missing", "observed")
)
ggplot(miss_map, aes(t, id, fill = status)) +
  geom_tile() +
  scale_fill_manual(values = c(observed = "grey85",
                               missing = weave_cols[["held_out"]]),
                    name = NULL) +
  scale_y_continuous(breaks = seq(5, n, by = 5)) +
  labs(x = "Week", y = "Site",
       title = "Where the gaps are",
       subtitle = "missing weeks (magenta) come in clustered runs, like real reporting dropouts") +
  theme_weave()
```

Here are the first four sites. The **navy line** is the true underlying mean
(\( \lambda = e^{\mu_s + f_{st}} \)); **grey points** are the counts we
actually observe; and **magenta points** are the *held-out truth* at the weeks
that went missing — what the model must reconstruct but never gets to see.

```{r truth-frames, echo = FALSE}
show     <- 1:4
df_truth <- data.frame(id = as.integer(true_data$id), t = true_data$t,
                       lambda = true_data$lambda, y = true_data$y)
df_obs   <- data.frame(id = as.integer(obs_data$id), t = obs_data$t,
                       y_obs = obs_data$y_obs)
df_miss  <- merge(df_truth, df_obs, by = c("id", "t"))
df_miss  <- df_miss[is.na(df_miss$y_obs), ]          # held-out truth

sub <- function(d) d[d$id %in% show, ]
```

```{r truth-plot, echo = FALSE, fig.height = 4.6}
ggplot() +
  geom_line(data = sub(df_truth), aes(t, lambda),
            colour = weave_cols[["truth"]], linewidth = 0.5) +
  geom_point(data = sub(df_obs), aes(t, y_obs),
             size = 0.7, colour = weave_cols[["observed"]]) +
  geom_point(data = sub(df_miss), aes(t, y),
             size = 0.9, colour = weave_cols[["held_out"]]) +
  facet_wrap(~ id, scales = "free_y", labeller = label_both) +
  labs(x = "Week", y = "Cases",
       title = "Simulated counts (4 of the sites)",
       subtitle = "navy = true mean · grey = observed · magenta = held-out truth") +
  theme_weave()
```

## 2. Fitting the kernel hyperparameters

Estimation happens in two moves: build a cheap stand-in for the latent field
we cannot see, then search for the kernel parameters under which that
stand-in looks most plausible. Missing weeks get a dedicated correction — an
EM-style refinement — described at the end of this part.

### A quick plug-in latent field

::: {.maths}
We cannot see the latent field \( f \) directly, so we build a cheap stand-in
\( g \) from the counts:

\[
g_{st} = \frac{\log(1 + y_{st}) - \bar{g}_s}{\text{sd}_s},
\]

that is, take \( \log(1+y) \), then for each site subtract that site's mean and
divide by its standard deviation.
:::

::: {.intuition}
The logarithm turns the multiplicative "rate" scale into an additive one that
matches \( f \). Subtracting each site's average removes the baseline
\( \mu_s \) (which we do not care about here), and dividing by the spread puts
every site on the same footing, so a single set of hyperparameters applies to
all of them. It is noisy — but it is instant.
:::

```{r plugin}
# the plug-in field: per-site centred & scaled log(1 + y), as one long vector
# (length n * nt, with time varying fastest)
g <- build_plugin_field(obs_data, n = n, nt = nt)
str(g)
```

For one site, the plug-in field tracks the true latent field, just with
observation noise layered on top:

```{r plugin-plot, echo = FALSE, fig.height = 3.6}
G        <- t(matrix(g, nrow = nt, ncol = n))        # n x nt
f_true_m <- t(matrix(log(true_data$lambda) - rep(coordinates$mu, each = nt),
                     nrow = nt, ncol = n))           # true f, centred per site

site <- 1
plot_df <- rbind(
  data.frame(t = 1:nt, value = scale(f_true_m[site, ])[, 1], type = "true f (scaled)"),
  data.frame(t = 1:nt, value = G[site, ],                    type = "plug-in g")
)
ggplot(plot_df, aes(t, value, colour = type)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("true f (scaled)" = weave_cols[["truth"]],
                                 "plug-in g" = weave_cols[["estimate"]]),
                      name = NULL) +
  labs(x = "Week", y = "Standardised latent field",
       title = "Site 1: plug-in field vs the truth",
       subtitle = "navy = true latent field · blue = plug-in estimate") +
  theme_weave()
```

### Scoring candidate hyperparameters: the marginal likelihood

::: {.maths}
For a candidate set of hyperparameters \( \theta \) the kernels give a
covariance \( K(\theta) \). We add a **nugget** — a noise term \( \eta \) —
and a global variance \( \sigma^2 \):

\[
g \sim \mathcal{N}\!\Big(0,\ \sigma^2\big[(R_{\text{space}} \otimes R_{\text{time}}) + \eta I\big]\Big),
\]

where \( R \) are *correlation* kernels (unit diagonal). Writing \( K \) for
this whole covariance and \( N = n \cdot n_t \) for the number of cells, the
score for \( \theta \) is the log marginal likelihood

\[
\log p(g \mid \theta) = -\tfrac{1}{2}\Big(\log|K| + g^\top K^{-1} g + N\log 2\pi\Big).
\]

We pick the \( \theta \) that maximises this. The variance \( \sigma^2 \) has a
closed-form optimum, so we "profile" it out and never search over it.

Strictly, the score also includes weakly informative log-normal priors on the
four parameters (`default_kernel_priors()`), so the estimate is a MAP rather
than a pure maximum likelihood — mild regularisation that keeps weakly
identified parameters (notably `long_term_scale`) off the boundary.
:::

::: {.intuition}
For any candidate hyperparameters we can ask: *how plausible is the plug-in
field if the world really had this much smoothness?* The first term
(\( \log|K| \)) penalises over-flexible kernels (an "Occam" penalty); the
second rewards kernels that explain the field well. We adjust the
hyperparameters until the score is best.

The **nugget** \( \eta \) is the important addition: a slice of pure noise the
model may attribute to observation error rather than signal. Without it the
kernel would have to explain the noisy plug-in field with smoothness alone,
badly distorting the length-scales.
:::

Evaluating \( \log|K| \) and \( g^\top K^{-1} g \) sounds hopeless — here
\( K \) is `r n * nt` square, and the optimiser scores thousands of
candidates. Because the kernel is separable, both quantities come exactly
from eigendecompositions of the two *small* kernels, at
\( O(n^3 + n_t^3) \) cost instead of \( O\big((n\,n_t)^3\big) \). How that
works is its own short story: see
[*The Kronecker trick*](kronecker.html).

### Handling the gaps: EM refinement

Filling the missing weeks with a flat per-site mean makes the field look as
if the signal dies in every gap, which biases the length-scales downward (the
seasonal and long-run scales suffer most). Setting `refine = TRUE` removes
that bias: the fit fills each gap with the GP's *own* conditional mean and
refits a handful of times.

::: {.intuition}
This loop is exactly the **EM algorithm** for a fit with missing data, alternating
two steps:

- **E-step.** With the current hyperparameters, replace each missing week by its
  expected value under the model — the GP **conditional mean** given the weeks we
  *did* observe (the same conditional-mean solve `gp_predict()` performs;
  "kriging", in geostatistics jargon).
- **M-step.** Treat that completed grid as if it were fully observed and
  re-maximise the fast (Kronecker) marginal likelihood to get updated
  hyperparameters.

Each pass feeds the next, and two or three are enough to converge — landing on
essentially the estimate we would have got with *no* gaps at all. Crucially, the
expensive conditional-mean solve runs only once per pass, never inside the
optimiser, so the whole correction stays cheap. (It removes the bias the *gaps*
cause; the residual plug-in attenuation listed in the caveats — from
conditioning on a noisy field rather than integrating it out — is a separate
matter it does not fix.)
:::

### The worked example: fit and check

`infer_kernel_params()` does all of the above: build the plug-in field, then
maximise the score over the three length-scales plus the nugget ratio, with
`refine = TRUE` switched on because our data has gaps.

```{r infer, warning = TRUE}
# maximise the GP marginal likelihood to recover the three length-scales
# (plus a noise/nugget ratio); the global variance sigma^2 is profiled out.
# refine = TRUE fills the gaps with the GP conditional mean and refits, removing
# the bias that mean-imputing the missing weeks would otherwise introduce.
est <- infer_kernel_params(obs_data, coordinates, nt = nt, period = period,
                           refine = TRUE)

# did we recover the hyperparameters we simulated from?
data.frame(
  parameter = c("length_scale", "periodic_scale", "long_term_scale"),
  truth     = c(true_length_scale, true_periodic_scale, true_long_term_scale),
  estimate  = round(c(est$length_scale, est$periodic_scale, est$long_term_scale), 2)
)
```

```{r infer-extras}
# the other two pieces estimation returns: the nugget and the profiled variance
cat(sprintf("nugget ratio (noise/signal) = %.2f;  profiled sigma^2 = %.2f\n",
            est$nugget_ratio, est$sigma2))

# est also carries the maximised score and the optimiser's convergence code
cat(sprintf("log posterior = %.1f;  convergence = %d (0 = success)\n",
            est$log_posterior, est$convergence))
```

::: {.caveat}
**At scale.** Estimation cost grows as \( O(n^3 + n_t^3) \). For very large
site counts, `n_sites` fits the hyperparameters on a random site subsample —
they are shared, population-level quantities, so a representative subsample
estimates the same values at a fraction of the cost (set a seed for
reproducibility; time points are never subsampled). Prediction cost is
dominated by the interval draws, which parallelise — see
[*Running predictions in parallel*](parallel.html).
:::

The clearest check is to draw the **fitted** kernels (blue) on top of the
**true** ones (navy, dashed). If the method worked, they overlap.

```{r kernel-check, echo = FALSE}
kernel_curves <- function(ls, ps, lts, label) {
  sp <- data.frame(distance = seq(0, 5, length.out = 200), type = label)
  sp$correlation <- rbf_kernel(sp$distance, theta = ls)
  tm <- data.frame(lag = seq(0, nt - 1, length.out = 400), type = label)
  tm$correlation <- periodic_kernel(tm$lag, alpha = ps, period = period) *
    rbf_kernel(tm$lag, theta = lts)
  list(space = sp, time = tm)
}
fit  <- kernel_curves(est$length_scale, est$periodic_scale, est$long_term_scale, "Estimated")
true <- kernel_curves(true_length_scale, true_periodic_scale, true_long_term_scale, "True")

cols <- c(Estimated = weave_cols[["estimate"]], True = weave_cols[["truth"]])
ltys <- c(Estimated = "solid",                  True = "dashed")

ggplot(rbind(fit$space, true$space),
       aes(distance, correlation, colour = type, linetype = type)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = cols, name = NULL) +
  scale_linetype_manual(values = ltys, name = NULL) +
  labs(x = "Spatial distance", y = "Correlation",
       title = "Spatial kernel: estimated vs true") +
  ylim(0, 1) + theme_weave()

ggplot(rbind(fit$time, true$time),
       aes(lag, correlation, colour = type, linetype = type)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = cols, name = NULL) +
  scale_linetype_manual(values = ltys, name = NULL) +
  labs(x = "Temporal lag (weeks)", y = "Correlation",
       title = "Temporal kernel: estimated vs true") +
  theme_weave()
```

## 3. From fitted kernels to predictions

With the kernels estimated, we put them to work: clean up the noise, fill in
the missing weeks, and say how *uncertain* each estimate is. A single
function, `gp_predict()`, does both of the pieces below in one call.

### The best guess: the posterior mean

::: {.maths}
`gp_predict()` conditions a separable Gaussian process on the **observed cells
only**. The posterior **mean** of the latent field comes from one matrix-free
conjugate-gradient solve (cheap because multiplying by \( K \) never forms it
— see [*The Kronecker trick*](kronecker.html)),

\[
\hat{f} = K S^\top \bigl(S K S^\top + \nu I\bigr)^{-1} g_{\text{obs}},
\]

where \( S \) simply picks out the observed cells and
\( \nu = \sigma^2 \eta \) is the observation-noise variance implied by the
fitted nugget ratio. The mean is smooth and deterministic — it does not
depend on how many draws are used for the interval.
:::

::: {.intuition}
Because it conditions on the observed set, missing weeks are filled by genuine
GP interpolation — the model's best guess given every week it *did* see,
weighted by the fitted correlations — unlike a smoother that quietly
mean-imputes the gaps.
:::

### The interval: exact variance plus a "gaps" correction

::: {.maths}
The posterior **variance** splits into two parts:

\[
\operatorname{Var}[f] \;=\;
  \underbrace{V_{\text{complete}}}_{\text{exact, closed form}}
  \;+\;
  \underbrace{\Delta_{\text{gaps}}}_{\text{estimated from } n_{\text{draws}} \text{ paired draws}} .
\]

If *no* weeks were missing, the posterior variance
\( V_{\text{complete}} = \operatorname{diag}\!\big(K - K(K+\nu I)^{-1}K\big) \)
has an exact closed-form answer through the same Kronecker eigendecomposition
used for fitting — no simulation needed. Missing weeks change that answer only
*near the gaps*, so the draws are spent purely on the correction
\( \Delta_{\text{gaps}} \). Each perturbation draw (Papandreou & Yuille 2010)
draws \( u \sim \mathcal{N}(0, K) \) and \( e \sim \mathcal{N}(0, \nu I) \)
and forms

\[
d_{\text{obs}} = u - K S^\top \bigl(S K S^\top + \nu I\bigr)^{-1}
  \bigl(S u + S e\bigr),
\]

which has exactly the posterior covariance of the field. It is paired with an
exact *complete-grid twin* \( d_{\text{twin}} \) built from the **same**
\( u, e \), and the average of \( d_{\text{obs}}^2 - d_{\text{twin}}^2 \)
estimates the correction. Sharing the randomness makes the difference nearly
noise-free — a **control variate**.
The latent-rate posterior is then combined with Negative-Binomial observation
noise, via the **law of total variance**, to give the count mean and variance

\[
\mathbb{E}[y] = \mathbb{E}[\lambda], \qquad
\operatorname{Var}[y] =
  \underbrace{\mathbb{E}\!\big[\lambda + \tfrac{\lambda^2}{r}\big]}_{\text{count noise}}
  + \underbrace{\operatorname{Var}[\lambda]}_{\text{rate uncertainty}} ,
\]

and the 95% interval comes from a **lognormal moment-match** to that mean and
variance:

\[
s^2 = \log\!\left(1 + \frac{\operatorname{Var}[y]}{\mathbb{E}[y]^2}\right),
\qquad
m = \log \mathbb{E}[y] - \tfrac{s^2}{2},
\]

with the interval given by the 2.5% and 97.5% quantiles of
\( \text{Lognormal}(m, s^2) \).
:::

::: {.intuition}
The variance trick is worth restating in plain words. Most of the uncertainty
is known **exactly** — the only thing simulation must measure is *how much the
gaps inflate it*. So instead of asking the draws to estimate the whole
variance (the old way, which needs hundreds of draws to average away the
noise), each draw is compared against an exact "what if nothing were missing"
twin that shares its random numbers. Almost all the randomness cancels in the
comparison — like judging a diet by weighing the same person before and after,
rather than comparing two different people. The result: weeks far from any gap
get an essentially exact interval, and a handful of draws does the work that
hundreds used to.

A future count is uncertain for two
reasons: we are unsure of the underlying rate, **and** counts scatter around
any given rate. An honest interval includes both. The dispersion \( r \) — how
overdispersed the counts are — is the one quantity this quick method does not
estimate from the likelihood, so it is recovered by method of moments from the
observed counts.
:::

### The worked example: predict, then check the coverage

The call returns one row per site-week with the posterior `rate` and, because
`n_draws >= 1`, a 95% interval (`lower`, `upper`). The dispersion `r` used is
attached as an attribute.

```{r predict, warning = TRUE}
# condition the GP on the observed cells and predict every site-week.
# n_draws controls the paired draws that estimate the gap correction to the
# variance -- the rest of the interval width is computed exactly, so modest
# values are plenty.
pred <- gp_predict(obs_data, coordinates, hyperparameters = est,
                   nt = nt, period = period, n_draws = 100)

head(pred)            # one row per cell: id, t, rate, and the 95% interval
attr(pred, "r")       # the NB dispersion used for the interval (estimated)
```

```{r pred-frames, echo = FALSE}
pred_df <- transform(pred, id = as.integer(id))   # integer ids for plotting/joins
```

The blue line is the predicted mean rate and the blue band the 95% prediction
interval. It should track the navy true-mean line and cover the points —
including the magenta held-out weeks it never saw.

```{r predict-plot, echo = FALSE, fig.height = 4.6}
ggplot() +
  geom_ribbon(data = sub(pred_df), aes(t, ymin = lower, ymax = upper),
              fill = weave_cols[["prediction"]], alpha = 0.18) +
  geom_line(data = sub(pred_df), aes(t, rate),
            colour = weave_cols[["prediction"]], linewidth = 0.7) +
  geom_line(data = sub(df_truth), aes(t, lambda),
            colour = weave_cols[["truth"]], linewidth = 0.4) +
  geom_point(data = sub(df_obs), aes(t, y_obs),
             size = 0.7, colour = weave_cols[["observed"]]) +
  geom_point(data = sub(df_miss), aes(t, y),
             size = 0.9, colour = weave_cols[["held_out"]]) +
  facet_wrap(~ id, scales = "free_y", labeller = label_both) +
  labs(x = "Week", y = "Cases",
       title = "Prediction vs truth",
       subtitle = "blue = predicted mean + 95% interval · navy = true mean · magenta = held-out truth") +
  theme_weave()
```

At scale, the draws behind the interval can run across CPU cores: see the
short companion article
[*Running predictions in parallel*](parallel.html) — one line of setup, and
the numbers are guaranteed not to change.

The honest test is the **held-out** weeks: does the 95% interval contain the
counts the model never saw? It should, about 95% of the time.

```{r diagnostics}
# keep only the held-out weeks (those the model never saw), with the true rate
held_out <- merge(df_miss[, c("id", "t", "lambda", "y")], pred_df,
                  by = c("id", "t"))

# coverage: fraction of held-out counts inside the 95% interval (target ~0.95)
mean(held_out$y >= held_out$lower & held_out$y <= held_out$upper)

# how well the predicted rate tracks the true rate at those weeks
cor(held_out$rate, held_out$lambda)
```

::: {.intuition}
If the coverage lands near 95% the bands are honest; if the predicted rate
correlates strongly with the true rate, the gap-filling is doing its job. Both
can drift a little here because this is a small example and the quick method
does not propagate every source of uncertainty (see below).
:::

## Caveats and approximations

::: {.caveat}
This is a fast, deterministic estimate — not a full posterior. The shortcuts
that make it quick also limit it:

- **Plug-in, not Bayesian.** We condition on a single noisy estimate of the
  latent field rather than integrating it out. This *attenuates* the
  length-scales (`periodic_scale` and `long_term_scale` tend to come out a
  little low). The nugget mitigates this but does not remove it.
- **`log(1 + y)`** is a crude stand-in for the latent log-rate, and is poorest
  at very low counts.
- **Per-site scaling** folds all per-site variance into a single global
  \( \sigma^2 \) and amplifies noise at low-count sites.
- **One homoscedastic nugget.** Real count noise is heteroscedastic (it grows
  with the rate); we summarise it with a single ratio \( \eta \).
- **Missing cells.** With `refine = TRUE` (used above) the fit fills the gaps
  with the GP conditional mean and refits, so missingness no longer biases the
  length-scales; the default `refine = FALSE` mean-imputes the gaps and is
  faster but attenuated. (`gp_predict()` always conditions on the observed cells
  only, so its interval widens over gaps either way.)
- **The dispersion \( r \)** is recovered by a rough method of moments, so the
  *width* of the count prediction interval inherits that error.
- **A point estimate.** We report the maximiser, with no uncertainty on the
  hyperparameters themselves.
:::

For a quick, honest summary of the spatial and temporal correlation structure —
or a sensible starting point for a fuller model — it does the job in about a
second.

## References

Papandreou, G. and Yuille, A. L. (2010). Gaussian sampling by local
perturbations. *Advances in Neural Information Processing Systems 23*.
